Replacing standard galaxy profiles with \\ mixtures of Gaussians

### 2012-09-19 documents/mixture_models.tex
citation; index typo
lupton profiles; bundy citation

% to-do
% -----
% - Brewer suggestion about KLD and ack Brewer
%   - need to discuss this and show results from it
% - decide which models will make it in
% - make useful figures
% - make figures showing use of the models in a real situation
% - make sure we are consistent about normalization! (do the profiles sum to unity or are they unity at \xi = 1?)
% - send to Jim Bosch and Kevin Bundy for comments

@@

\newcommand{\tmatrix}[1]{\boldsymbol{#1}}
\newcommand{\inverse}[1]{{#1}^{-1}}
\newcommand{\transpose}[1]{{#1}^{\mathsf T}}
\newcommand{\tvector}[1]{\boldsymbol{#1}}
\newcommand{\pos}{\tvector{x}}
\newcommand{\spos}{\tvector{\xi}}
\newcommand{\mean}{\tvector{m}}
\newcommand{\var}{\tmatrix{V}\!}
\newcommand{\Gm}{\tmatrix{G}}
\newcommand{\Hm}{\tmatrix{H}}
\newcommand{\affine}{\tmatrix{R}}
\newcommand{\uv}{\tvector{u}}
\newcommand{\zero}{\tmatrix{0}}
\newcommand{\identity}{\tmatrix{I}}
\newcommand{\normal}{N}
\newcommand{\given}{\,|\,}
\renewcommand{\star}{\mathrm{star}}
\newcommand{\dev}{\mathrm{dev}}

@@

We need good performance in two-dimensional image synthesis (for
fitting) because, with \thetractor\ (Lang \etal, forthcoming), we are
building a comprehensive model of all the imaging data we can; this
will have models of many millions of galaxies in imaging that contains
on the order of $10^{13}$ pixels.  We are basing our models on the
\sdss\ Catalog galaxy models, which only include exponential,
de~Vaucouleurs, and composite (mixture of the two) radial profiles.
In detail, in fact, the \sdss\ Catalog models are small modifications
of these (\citealt{lupton}); details below.  So our goal here is to
provide replacements for these models in order to improve the
performance of galaxy image modeling and analysis software.  Some of
the models we present have been used previously in the study of XXX
(\cite{bundy}); they could also be used to speed other image-fitting
systems, like the very successful \project{GALFIT} (\citealt{galfit}).

@@

We wish to make an approximation $Q_{\exp}(\cdot)$ to the
two-dimensional circular exponential profile
\begin{eqnarray}\displaystyle
Q_{\exp}(\spos) &\approx& \exp(-\alpha_{\exp}\,|\spos|)
\\
\alpha_{\exp} &=& -1.67834699
\quad ,
\end{eqnarray}
where $\spos$ is a dimensionless focal-plane position, and
$\alpha_{\exp}$ ensures that the profile has unit half-light radius.
The position $\spos$ is dimensionless because it parameterizes the
unit-size and unit-flux dimensionless function.  We seek the best
(where ``best'' will be defined below) $M_g$-Gaussian mixture
approximation
\begin{eqnarray}\displaystyle
Q_{\exp}(\spos) &=& \sum_{m=1}^{M_g} g_m\,\normal(\spos\given\zero,\Gm_m)
\\
\Gm_m &=& \sigma^2_{\exp,m}\,\identity
\quad ,
\end{eqnarray}
where all of the means are exactly zero and all of the variances
$\Gm_m$ in the mixture are scalar multiples of the identity matrix
$\identity$ because we are requiring this dimensionless function to be
precisely circular (so every component is itself circular and
concentric).  Similarly for the de~Vaucouleurs profile
\begin{eqnarray}\displaystyle
Q_{\dev}(\spos) &\approx& \frac{1}{Z_{\dev}}\,\exp(-\alpha_{\dev}\,|\spos|^{1/4})
\\
\alpha_{\dev} &=& -7.66924944
\\
Q_{\dev}(\spos) &=& \sum_{m=1}^{M_h} h_m\,\normal(\spos\given\zero,\Hm_m)
\\
\Hm_m &=& \sigma^2_{\dev,m}\,\identity
\quad .
\end{eqnarray}
...Cite (\citealt{ciotti}) re half-light radius constants...

The Lupton profiles are...

...Lupton profiles as different from de~Vaucouleurs...Sersic profiles...Discuss the cutoffs and regularization used by PHOTO and adjust $Z$ and $r$ accordingly...

@@

Applying this PSF to an exponential or de~Vaucouleurs galaxy is
slightly more complicated, because the galaxy has not just a flux
$S_g$ and a central position $\pos_g$; it also has a shape.  Because
we are only considering these simple galaxies, we are only permitting
ellipsoidal shapes, which can be represented by a semi-major axis $a$,
a semi-minor axis $b$, and a position angle $\phi$, or equivalently by
eigenvalues $a, b$ and eigenvectors $\uv_1, \uv_2$, or by a symmetric
variance tensor $\var_g$ (which is a general representation of an
ellipse).  The galaxy is distorted by this variance tensor
\emph{prior} to PSF convolution, so the focal-plane image
(PSF-convolved intensity field) for a general exponential galaxy is
given by
\begin{eqnarray}\displaystyle
I(\pos\given\exp,S_g,\pos_g,\var_g) &=& \sum_{k=1}^K \sum_{m=1}^{M_g} S_g\,g_m\,p_k\,\normal(\pos\given\pos_g+\mean_k,\var_{gn}+\var_k)
\\
\var_{gn} &\equiv& \affine_g\cdot\Gm_m\cdot\transpose{\affine_g}
\\
\var_g &\equiv& \affine_g\cdot\transpose{\affine_g}
\\
\affine_g &=& \left[a\,\uv_1 , b\,\uv_2 \right]
\quad ,
\end{eqnarray}
where $a$ and $b$ are the major and minor axis lengths of the galaxy
ellipse (in appropriate units) and $\uv_1$ and $\uv_2$ are the
eigenvectors on the sky pointing in the major-axis and minor-axis
directions respectively.  Since implicitly all vectors are
two-dimensional column vectors, this makes $\affine_g$ a $2\times 2$
affine transformation matrix and $\var_g$ a symmetric $2\times 2$
variance tensor.  Similarly, the general de~Vaucouleurs galaxy is
given by
\begin{eqnarray}\displaystyle
I(\pos\given\dev,S_h,\pos_h,\var_h) &=& \sum_{k=1}^K \sum_{m=1}^{M_h} S_h\,h_m\,p_k\,\normal(\pos\given\pos_h+\mean_k,\var_{hn}+\var_k)
\\
\var_{hn} &\equiv& \affine_h\cdot\Hm_m\cdot\transpose{\affine_h}
\\
\var_h &\equiv& \affine_h\cdot\transpose{\affine_h}
\quad ,
\end{eqnarray}
where the matrix mathematics is the same as that in the exponential
case.  Note the important and key result of this \documentname, to
wit, that a mixture-of-Gaussian galaxy model (with $M$ components)
convolved with a mixture-of-Gaussian PSF model (with $K$ components)
yields a mixture-of-Gaussian image (with $[M\,K]$ components).

>>>>

% to-do
% -----
% - Brewer suggestion about KLD and ack Brewer
%   - need to discuss this and show results from it
% - decide which models will make it in
% - make useful figures
% - make figures showing use of the models in a real situation
% - make sure we are consistent about normalization! (do the profiles sum to unity or are they unity at \xi = 1?)
% - does the Lupton paper give the profiles, or should we cite the code directly?
% - send to Jim Bosch and Kevin Bundy for comments

@@

\newcommand{\tmatrix}[1]{\boldsymbol{#1}}
\newcommand{\inverse}[1]{{#1}^{-1}}
\newcommand{\transpose}[1]{{#1}^{\mathsf T}}
\newcommand{\tvector}[1]{\boldsymbol{#1}}
\newcommand{\pos}{\tvector{x}}
\newcommand{\spos}{\tvector{\xi}}
\newcommand{\mean}{\tvector{m}}
\newcommand{\var}{\tmatrix{V}\!}
\newcommand{\affine}{\tmatrix{R}}
\newcommand{\uv}{\tvector{u}}
\newcommand{\zero}{\tmatrix{0}}
\newcommand{\identity}{\tmatrix{I}}
\newcommand{\normal}{N}
\newcommand{\given}{\,|\,}
\renewcommand{\star}{\mathrm{star}}
\newcommand{\dev}{\mathrm{dev}}
\newcommand{\lux}{\mathrm{lux}}
\newcommand{\luv}{\mathrm{luv}}

@@

We need good performance in two-dimensional image synthesis (for
fitting) because, with \thetractor\ (Lang \etal, forthcoming), we are
building a comprehensive model of all the imaging data we can; this
will have models of many millions of galaxies in imaging that contains
on the order of $10^{13}$ pixels.  We are basing our models on the
\sdss\ Catalog galaxy models, which only include exponential,
de~Vaucouleurs, and composite (mixture of the two) radial profiles.
In detail, in fact, the \sdss\ Catalog models are small modifications
of these (\citealt{lupton}); details below.  So our goal here is to
provide replacements for these models in order to improve the
performance of galaxy image modeling and analysis software.  Some of
the models we present have been used previously in tools for precise
photometry (Bundy \etal, forthcoming); they could also be used to
speed other image-fitting systems, like the very successful
\project{GALFIT} (\citealt{galfit}).

@@

We wish to make an approximation to the two-dimensional circular
exponential profile $Q_{\exp}(\cdot)$
\begin{eqnarray}\displaystyle
Q_{\exp}(\spos) &\equiv& \exp(-\alpha_{\exp}\,[|\spos| - 1])
\\
\alpha_{\exp} &\equiv& -1.67834699
\quad ,
\end{eqnarray}
where $\spos$ is a dimensionless focal-plane position, and
$\alpha_{\exp}$ ensures that the profile has unit half-light radius.
The position $\spos$ is dimensionless because it parameterizes the
unit-size and unit-flux dimensionless function.  We seek the best
(where ``best'' will be defined below) $M_g$-Gaussian mixture
approximation
\begin{eqnarray}\displaystyle
Q_{\exp}(\spos) &\approx& \sum_{m=1}^{M_{\exp}} a_{\exp,m}\,\normal(\spos\given\zero,\var_{\exp,m})
\\
\var_{\exp,m} &\equiv& \sigma^2_{\exp,m}\,\identity
\quad ,
\end{eqnarray}
where all of the means are exactly zero and all of the variances
$\var_m$ in the mixture are scalar multiples of the identity matrix
$\identity$ because we are requiring this dimensionless function to be
precisely circular (so every component is itself circular and
concentric).  Similarly for the de~Vaucouleurs profile
\begin{eqnarray}\displaystyle
Q_{\dev}(\spos) &\equiv& \exp(-\alpha_{\dev}\,[|\spos|^{1/4} - 1])
\\
\alpha_{\dev} &\equiv& -7.66924944
\\
Q_{\dev}(\spos) &\approx& \sum_{m=1}^{M_{\dev}} a_{\dev,m}\,\normal(\spos\given\zero,\var_{\dev,m})
\\
\var_{\dev,m} &\equiv& \sigma^2_{\dev,m}\,\identity
\quad .
\end{eqnarray}
(The half-light radius parameters $\alpha_{\exp}$ and $\alpha_{\dev}$
are taken from \citealt{ciotti}.)  The challenge we meet below is to
determine the parameters
\begin{eqnarray}
\{a_{\exp,m},\sigma_{\exp,m}\}_{m=1}^{M_{\exp}} \mbox{~and~} \{a_{\dev,m},\sigma_{\dev,m}\}_{m=1}^{M_{\dev}}
\end{eqnarray}
to best approximate the
traditional galaxy profile functions, under some sensible definiton of
the word ``best'', as a function of the model complexity parameters
(numbers of components) $(M_{\exp}, M_{\dev})$.

The \sdss\ pipelines make use of modified profiles, which have been
truncated smoothly at large radius and (in the case of the
de~Vaucouleurs profile) ``softened'' at the center (\citealt{lupton}).
The \sdss\ form of the exponential profile is
\begin{eqnarray}\displaystyle
Q_{\lux}(\spos) &\equiv& \left\{\begin{array}{ll}
  \exp(-\alpha_{\lux}\,[|\spos| - 1]) & \mbox{for~}|\spos| < 3 \\
  \exp(-\alpha_{\lux}\,[|\spos| - 1])
  \,\left[1 - [|\spos| - 3]^2\right]^2 & \mbox{for~}3 < |\spos| < 4 \\
  0                                   & \mbox{for~}4 < |\spos|
\end{array}\right.
\\
\alpha_{\lux} &\equiv& -1.67835
\quad ,
\end{eqnarray}
and the \sdss\ form of the de~Vaucouleurs profile is
\begin{eqnarray}\displaystyle
Q_{\luv}(\spos) &\equiv& \left\{\begin{array}{ll}
  \exp(-\alpha_{\luv}\,\left[[|\spos|^2 + 0.0004]^{1/8} - 1\right]) & \mbox{for~}|\spos| < 7 \\
  \exp(-\alpha_{\luv}\,\left[[|\spos|^2 + 0.0004]^{1/8} - 1\right])
  \,\left[1 - [|\spos| - 7]^2\right]^2 & \mbox{for~}7 < |\spos| < 8 \\
  0                                   & \mbox{for~}8 < |\spos|
\end{array}\right.
\\
\alpha_{\luv} &\equiv& -7.66925
\quad .
\end{eqnarray}
(The half-light radius parameters $\alpha_{\exp}$ and
$\alpha_{\dev}$---and the softening and cutoff radius parameters---are
taken from the \sdss\ codebase.)  The \sdss\ radial profiles are much
better behaved than the standard forms; we recommend them strongly.
Once again, the challenge is to determine the best parameters
\begin{eqnarray}
\{a_{\lux,m},\sigma_{\lux,m}\}_{m=1}^{M_{\lux}} \mbox{~and~} \{a_{\luv,m},\sigma_{\luv,m}\}_{m=1}^{M_{\luv}}
\quad .
\end{eqnarray}

@@

Applying this PSF to an exponential or de~Vaucouleurs galaxy is
slightly more complicated, because the galaxy has not just a flux
$S_g$ and a central position $\pos_g$; it also has a shape.  Because
we are only considering these simple galaxies, we are only permitting
ellipsoidal shapes, which can be represented by a semi-major axis $a$,
a semi-minor axis $b$, and a position angle $\phi$, or equivalently by
eigenvalues $a, b$ and eigenvectors $\uv_1, \uv_2$, or by a symmetric
variance tensor $\var_g$ (which is a general representation of an
ellipse).  The galaxy is distorted by this variance tensor
\emph{prior} to PSF convolution, so the focal-plane image
(PSF-convolved intensity field) for a general exponential galaxy is
given by
\begin{eqnarray}\displaystyle
I(\pos\given\exp,S_g,\pos_g,\var_g) &=& \sum_{k=1}^K \sum_{m=1}^{M_g} S_g\,g_m\,p_k\,\normal(\pos\given\pos_g+\mean_k,\var_{gm}+\var_k)
\\
\var_{gm} &\equiv& \affine_g\cdot\var_m\cdot\transpose{\affine_g}
\\
\var_g &\equiv& \affine_g\cdot\transpose{\affine_g}
\\
\affine_g &=& \left[a\,\uv_1 , b\,\uv_2 \right]
\quad ,
\end{eqnarray}
where $a$ and $b$ are the major and minor axis lengths of the galaxy
ellipse (in appropriate units) and $\uv_1$ and $\uv_2$ are the
eigenvectors on the sky pointing in the major-axis and minor-axis
directions respectively.  Since implicitly all vectors are
two-dimensional column vectors, this makes $\affine_g$ a $2\times 2$
affine transformation matrix and $\var_g$ a symmetric $2\times 2$
variance tensor.  Similarly, the general de~Vaucouleurs galaxy is
given by
\begin{eqnarray}\displaystyle
I(\pos\given\dev,S_h,\pos_h,\var_h) &=& \sum_{k=1}^K \sum_{m=1}^{M_h} S_h\,h_m\,p_k\,\normal(\pos\given\pos_h+\mean_k,\var_{hn}+\var_k)
\\
\var_{hn} &\equiv& \affine_h\cdot\var_m\cdot\transpose{\affine_h}
\\
\var_h &\equiv& \affine_h\cdot\transpose{\affine_h}
\quad ,
\end{eqnarray}
where the matrix mathematics is the same as that in the exponential
case.  Note the important and key result of this \documentname, to
wit, that a mixture-of-Gaussian galaxy model (with $M$ components)
convolved with a mixture-of-Gaussian PSF model (with $K$ components)
yields a mixture-of-Gaussian image (with $[M\,K]$ components).
